Investigating Determinism of Gaming Engines for Autonomous Vehicle Verification

### 2019-12-19 paper/1_Intro.tex
Added discussed points and edited structure a bit, need to ask greg about some of the changes he did

% What is verification
\IEEEPARstart{V}{}erification is the process used to gain confidence in the correctness of a system with respect to its requirements~\cite{bergeron2012writing}. Simulation based verification can be used to test autonomous driving functions under development benefiting from full control over the road network and the actors within it. These simulated tests aim to provide evidence to regulators of the functional safety of the vehicle or its compliance with commonly agreed upon road conduct~\cite{ViennaConv}, national rules~\cite{codes2015highway} and road traffic laws~\cite{RoadTraffic1988}. 
%
\noindent There have been a number of fatalities with autonomous vehicles which could be attributed to a lack of verification and validation (V\&V), e.g.\cite{FatalityExample}.
%
Simulation is an obvious domain choice to explore the vast parameter space in a safe and efficient manner~\cite{korosec2019waymo} without millions of miles of costly on-road testing~\cite{kalra2016driving} where the rare event can be made to happen more often~\cite{Koopman2018} such as unpredictable actor behaviour~\cite{RobustnessAutonomy}. 
%

% Requirements of simulation based verification SBV
For games engines to offer a suitable testing environment they must: operate with a minimal \textit{reality gap} in the physical domain (vehicles, actors), be realistic enough \cite{Koopman2018} including suitable rendering for perception stack testing, be easy to setup, run and manipulate the environment and the temporal development of actors \cite{Ulbrich2015}, include the provision for sensor analogues to a suitable accuracy (e.g. video, LIDAR, IR), allow hardware-in-the-loop development and provide a suitable real-time test-bed for cyber-security testing~\cite{Javaid2013}. 
%, e.g. minimum rate of frames per second (fps).
%
% Game engines for SBV
Game engines offer a simulation domain solution for the development and testing of CAVs that meet many of these requirements but many challenges still exist.
%
% Verification and Determinism 
For games engines to be useful as a verification tool then they must be deterministic. 
%
% What is a deterministic system
Deterministic, in this context, refers to the property of causality given a temporal development of events such that any state is completely determined by prior states.
%
Herein the term \textit{simulation variance} is used to refer to how simulation runs vary when repeated many times and specifically the variance of agent paths within a given scenario. We will adopt the terminology defined in~\cite{Ulbrich2015}, where \textit{scene} refers to all static objects including the road network, street furniture, environment conditions and a snapshot of any dynamic elements. Dynamic elements are the elements in a scene whose actions or behaviour may change over time, these are considered actors and may include other road vehicles, cyclists, pedestrians and traffic signals. The \textit{scenario} is then defined as a temporal development between several scenes which may be specified by specific goals and values. A \textit{situation} is defined as the subjective conditions and determinants for behaviour at a particular point in time.
% 
% Why is determinism important for verification
Determinism is important for verification. Any software defects that are detected but not repeatable at a later time or by another party does not allow for suitable resolution. Furthermore, defects not found on the verification testing system may arise on a deployed system with a different hardware architecture resulting in untested and potentially unsafe vehicle behaviour.

>>>>

% What is verification
\IEEEPARstart{V}{}erification is the process used to gain confidence in the correctness of a system with respect to its requirements~\cite{bergeron2012writing}. Simulation based verification can be used to test autonomous driving functions under development benefiting from full control over the road network and the actors within it. These simulated tests aim to provide evidence to regulators of the functional safety of the vehicle or its compliance with commonly agreed upon road conduct~\cite{ViennaConv}, national rules~\cite{codes2015highway} and road traffic laws~\cite{RoadTraffic1988}. 
%
There have been a number of fatalities with autonomous vehicles which could be attributed to a lack of verification and validation (V\&V), e.g.\cite{FatalityExample}.
%
Simulation is an obvious domain choice to explore the vast parameter space in a safe and efficient manner~\cite{korosec2019waymo} without millions of miles of costly on-road testing~\cite{kalra2016driving} where the rare event can be made to happen more often~\cite{Koopman2018} such as unpredictable actor behaviour~\cite{RobustnessAutonomy}.

% Requirements of simulation based verification SBV
For games engines to offer a suitable testing environment they must: operate with a minimal \textit{reality gap} in the physical domain (vehicles, actors), be realistic enough \cite{Koopman2018} including suitable rendering for perception stack testing, be easy to setup, run and manipulate the environment and the temporal development of actors \cite{Ulbrich2015}, include the provision for sensor analogues to a suitable accuracy (e.g. video, LIDAR, IR), allow hardware-in-the-loop development and provide a suitable real-time test-bed for cyber-security testing~\cite{Javaid2013}. 
%, e.g. minimum rate of frames per second (fps).
%
% Game engines for SBV
Game engines offer a simulation domain solution for the development and testing of CAVs that meet many of these requirements, as opposed to vehicle dynamics simulators and traffic-level simulators used my manufacturers~\cite{FrameworkAndChallenges}; but many challenges still exist.

% Verification and Determinism 
For game engines to be useful as a verification tool then they must be deterministic. 
%
% What is a deterministic system
Deterministic, in this context, refers to the property of causality given a temporal development of events such that any state is completely determined by prior states.
%
Herein the term \textit{simulation variance} is used to refer to how simulation runs vary when repeated many times and specifically the variance of agent paths within a given scenario. We will adopt the terminology defined in~\cite{Ulbrich2015}, where \textit{scene} refers to all static objects including the road network, street furniture, environment conditions and a snapshot of any dynamic elements. Dynamic elements are the elements in a scene whose actions or behaviour may change over time, these are considered actors and may include other road vehicles, cyclists, pedestrians and traffic signals. The \textit{scenario} is then defined as a temporal development between several scenes which may be specified by specific goals and values. A \textit{situation} is defined as the subjective conditions and determinants for behaviour at a particular point in time.
% 
% Why is determinism important for verification
Determinism is important for verification. Any software defects that are detected but not repeatable at a later time or by another party does not allow for suitable resolution. Furthermore, defects not found on the verification testing system may arise on a deployed system with a different hardware architecture resulting in untested and potentially unsafe vehicle behaviour.

### 2019-12-19 paper/2_Background.tex
Added discussed points and edited structure a bit, need to ask greg about some of the changes he did

There are numerous game engines with their associated development environments which could be suitable for CAV development, e.g. Unreal, Unity, CryEngine. Specific autonomous driving research tools have been created to abstract and simplify the development environment, some of which are based on existing engines, e.g. Carla, AirSim, Apollo, and some have been developed specifically, e.g. the could based Nvidia Drive Constellation simulator. \TODO{ref each of these}
%
Investigating the determinism of gaming engines has not attracted much research interest since performance is more critical for playing games than accurate and repeatable runtime execution. However, when these engines are used for CAV development then deterministic behaviour is required. This section gives an overview of a gaming engine and what sources or settings in the engine may affect \textit{simulation variance}.
%
% \noindent This section overviews important information and concepts in physics engines, which are relevant to their usage in CAV simulations. Particularly, on how physics engines operate and the sources of non-determinism in these engines necessary to understand the contribution of this paper.

% ======= Overview of a Game Loop
\subsection{Overview of a Game Loop} \label{GameLoopSection}
The game loop is responsible for the interaction between the maths, physics and rendering engines~\cite{GameEngineArchBook}. Figure \ref{GameEngineLoopDiagram} depicts a basic representation of the process flow in a game engine loop. A traditional \TODO{For AG: how is it traditional, what is the alternative?} game loop is broken up into three distinct phases: processing the inputs, updating the game world (Physics Engine), and generating outputs (Rendering).

\begin{figure}[h]
\centering
\includegraphics[width=0.5\textwidth]{Other/Figures/GameEngineLoop.pdf}
\caption{Game engine loop block diagram}
\label{GameEngineLoopDiagram}
\end{figure}
The first part of the game loop is to process the user inputs which may take the form of user keyed inputs or, in the case of CAV development, the resultant actions of the autonomous vehicle given the current state of the environment. \TODO{For AG: need to expand, e.g what are the inputs?} \TODO{For AG: does game logic need to be mentioned?}
%
The update interprets the intended actions of the AV along with all other dynamic actors in the scene and performs any necessary physics calculations and updates the actor states. \TODO{For AG: what does the clock symbolise, need to explain it?}
%
The rendering engine then takes the updated actor states and renders the scene~\cite{GameProgPatternsBook}. \TODO{For AG: you cite this book 4 times which might be too much as a general reference, is there something specific you want to draw the reader towards in the book? What specifically about the rendering part of this cycle does the book help to illustrate the point?}

It is crucial for this paper to understand how the rendering and physics engines operate together. \TODO{For AG: why is this crucial and can we say that we resolved it if it is crucial?}
Briefly, the game loop operates as follows:\cite{GameProgPatternsBook} (see Figure \ref{GameEngineLoopDiagram})
\begin{itemize}[leftmargin=*]
    \item At the beginning of each tick (scene), the lag between the game clock and the real world is updated based on how much real time passed. This measures how far the game's clock is behind compared to the real world.
    \item Then the user inputs are processed.
    \item There is then an inner loop to update the physics engine (clock symbol in Fig.~\ref{GameEngineLoopDiagram}), incrementing at fixed step until the game clock is equal to the real world. The physics engine uses a fixed time step, because it makes everything simpler and more stable for physics and AI. The shorter this fixed time step is, the more processing time it takes to catch up to real time and the more deterministic the engine becomes and vice versa\cite{GameEngineArchBook}\cite{GameProgPatternsBook}, as will be explained in section \ref{s:nondeterminisimSources}. Thus, ideally the time step should be as short as possible, so that simulations run with high fidelity on fast machines. However, if the fixed time step is too short i.e. less than the time it takes to process an ``Update" inner loop on some (slow) machines, then the simulation will simply never catch up on these slow machines.
    \item Rendering occurs once the physics engine catches up. The process then starts again.
\end{itemize}

The time step, $dt$, in any physics calculation is important. To use a fixed physics time step the user's display refresh rate needs to be known in advance. This requires an update loop to take less than one frame of real world time~\cite{gaffer}. Given the different capabilities of game players hardware a variable delta time can be implemented taking the previous frame time as the next $dt$. However, variable $dt$ can lead to different physical results and in some cases unrealistic physics. Semi-fixed or limited frame rates ensure $dt$ does not exceed some user-defined limit to meet a minimum standard of physical representation but allows computational headroom for slower hardware. Some engines provide sub-stepping which does multiple physics calculations per frame at a greater CPU cost~\footnote{https://docs.unrealengine.com/en-US/Engine/Physics/Substepping/index.html}. If the engine tries to render between updates \textit{residual lag} can occur \cite{GameEngineArchBook}\cite{GameProgPatternsBook} and extrapolation between frames is performed to smooth transition between scenes.

@@

% ======= Sources of non-determinism
\subsection{Sources of non-determinism} \label{s:nondeterminisimSources}
\noindent After establishing an understanding of how physics engines work, the focus will now be on what causes these engines to be non-deterministic. 
The main reasons that could cause non-determinism are discussed below. \\\\

\noindent\underline{\textit{Floating Points:}}
A generic issue with computers are floating points precision. 
Various errors occur when doing arithmetic manipulations using floating points, especially when operating between large and small numbers due to rounding, memory limitations and so forth.\cite{FloatingPointArithmeticArticle}\cite{FloatingPointsBook}. 
Translating to the usage of CAV simulations for V\&V in physics engines, this could result in many precision issues resulting in complex non deterministic behaviours.\\\\ 
\noindent For instance, consider a simulation where the world is infinite and progressively  generated in the physics engine. 
After a few hundred meters, precision issues will start to occur, and will get progressively worse the further from the origin the simulation gets. 
This is due to the values of the coordinates being used in calculations are getting bigger and hence will not cope with small values that are critical for V\&V testing purposes. 
A solution to this problem is perhaps every time the simulations moves 100 meters away from the origin the entire world will move by 100m in the opposite direction. Thus, avoiding getting into floating point issues.\cite{FloatingPointArithmeticArticle}\\\\
\noindent There are many scenarios that could be addressed, and solutions to them will entirely depend on the characteristics of that specific scenario. There is no generalised solution for the floating problem, regardless of how good computers get there will always be a limitation. However, floating points as they stand could just be sufficient for V\&V testing purposes. Whenever they are not, smart solutions can be found to accommodate for these limitations.\\\\
\begin{figure*}[b]
    \centering
    \includegraphics[width=0.99\linewidth]{Other/Figures/MethodologyDiagramCropped.pdf}
    \caption{Shows a flow diagram of the methodology proposed to define the level of non-determinism of a physics engine.}

    \label{method_diagram}
\end{figure*}
\noindent\underline{\textit{Navigation meshes and AI:}}
Some CAV simulator developers claim that the built-in physics engines' AI is non-deterministic\cite{CARLABenchmark}. 
The validity of this depends on what are the built in AI algorithms being used by the engine. 
Most of them tend to use the A* algorithm\cite{AStarBook}, which is an algorithm with deterministic behaviour if the environment is deterministic\cite{AirsimUnrealArticle}\cite{UnrealAIDocumentation}. 
Therefore the determinism of the AI depends on the determinism of the engine and how its navigation meshes are created or modelled. It is interesting to note that changes can occur to meshes every time the simulation is loaded. This is mainly resorted to the different scheduling of threads. \\\\
\noindent\underline{\textit{Physics and Rendering clocks:}}
The inherent way of how these physics engines operate (i.e.game loops) causes the engine to be non-deterministic. 
A follow up from section \ref{GameLoopSection}, these engines use a variable frame rate, which is good for hardware scalability but creates a challenge for the physics engine which works best with small fixed time steps. 
The problem of having a variable frame rate could also be a reason for the non-deterministic behaviour of these engines. This issue could perhaps be of negligible significance if the time steps are small enough. \\

\noindent\underline{\textit{Scheduling of threads:}} Scheduling of threads is believed to be the main reason for the non deterministic behaviour of these engines. Given the same hardware the output should always be the same, unless scheduling of the threads change, this would then cause non-deterministic behaviours. To solve this one would have to control all of the threads, which is a very involved and to achieve that one would need to replace or control the whole run-time system to allocate the same threads in the same order.

>>>>

There are numerous game engines with their associated development environments which could be suitable for CAV development, e.g. Unreal, Unity, CryEngine. Specific autonomous driving research tools have been created to abstract and simplify the development environment, some of which are based on existing engines, e.g. Carla, AirSim, Apollo, and some have been developed specifically, e.g. the could based Nvidia Drive Constellation simulator. \TODO{ref each of these}

Investigating the determinism of gaming engines has not attracted much research interest since performance is more critical for playing games than accurate and repeatable runtime execution. However, when these engines are used for CAV development then deterministic behaviour is required. This section gives an overview of a gaming engine and what sources or settings in the engine may affect \textit{simulation variance}.
%
% \noindent This section overviews important information and concepts in physics engines, which are relevant to their usage in CAV simulations. Particularly, on how physics engines operate and the sources of non-determinism in these engines necessary to understand the contribution of this paper.

% ======= Overview of a Game Loop
\subsection{Overview of a Game Loop} \label{GameLoopSection}
The game loop is responsible for the interaction between the maths, physics and rendering engines. Figure \ref{GameEngineLoopDiagram} depicts a basic representation of the process flow in a game engine loop. A game loop is broken up into three distinct phases: processing the inputs, updating the game world (Physics Engine), and generating outputs (Rendering).~\cite{GameEngineArchBook}

\begin{figure}[h]
\centering
\includegraphics[width=0.5\textwidth]{Other/Figures/GameEngineLoop.pdf}
\caption{Game engine loop block diagram}
\label{GameEngineLoopDiagram}
\end{figure}
The first part of the game loop is to process the user inputs which may take the form of user keyed inputs or, in the case of CAV development, the resultant actions of the autonomous vehicle given the current state of the environment. For example, the throttle and braking inputs from the PID controller controlling the speed of an AV. %\TODO{For AG: need to expand, e.g what are the inputs?} %\TODO{For AG: does game logic need to be mentioned?}
%
The update interprets the intended actions of all dynamic actors in the scene, performs any necessary physics calculations and updates the actor states. %This update process can still proceed independelty of the main game loop as symbolised by the clock in the upper loop in Figure~\ref{GameEngineLoopDiagram}. % \TODO{For AG: what does the clock symbolise, need to explain it?}
%
The rendering engine then takes the updated actor states and renders the scene~\cite{GameProgPatternsBook}. %\TODO{For AG: you cite this book 4 times which might be too much as a general reference, is there something specific you want to draw the reader towards in the book? What specifically about the rendering part of this cycle does the book help to illustrate the point?}

It is necessary to have an understanding of how rendering and physics engines operate together in order to grasp some of the sources of non-determinsim that will be mentioned. %\TODO{For AG: why is this crucial and can we say that we resolved it if it is crucial?}
The game loop operates as follows: (see Figure \ref{GameEngineLoopDiagram})
\begin{itemize}[leftmargin=*]
    \item At the beginning of each tick (scene), the lag between the game clock and the real world is updated based on how much real time passed. This measures how far the game's clock is behind compared to the real world.
    \item Then the user inputs are processed.
    \item There is then an inner loop to update the physics engine (clock symbol in Figure~\ref{GameEngineLoopDiagram}), incrementing at fixed step until the game clock is equal to the real world. The physics engine uses a fixed time step, because it makes everything simpler and more stable for physics and AI. The shorter this fixed time step is, the more processing time it takes to catch up to real time and the more deterministic the engine becomes and vice versa, as will be explained in section \ref{s:nondeterminisimSources}. Thus, ideally the time step should be as short as possible, so that simulations run with high fidelity on fast machines. However, if the fixed time step is too short i.e. less than the time it takes to process an ``Update" inner loop on some (slow) machines, then the simulation will simply never catch up on these slow machines.
    \item Rendering occurs once the physics engine catches up. The process then starts again.
\end{itemize}

The time step, $dt$, in any physics calculation is important. To use a fixed physics time step the user's display refresh rate needs to be known in advance. This requires an update loop to take less than one frame of real world time~\cite{gaffer}. Given the different capabilities of game players hardware a variable delta time can be implemented taking the previous frame time as the next $dt$. However, variable $dt$ can lead to different physical results and in some cases unrealistic physics. Semi-fixed or limited frame rates ensure $dt$ does not exceed some user-defined limit to meet a minimum standard of physical representation but allows computational headroom for slower hardware. Some engines provide sub-stepping which does multiple physics calculations per frame at a greater CPU cost~\footnote{https://docs.unrealengine.com/en-US/Engine/Physics/Substepping/index.html}. If the engine tries to render between updates \textit{residual lag} can occur and extrapolation between frames is performed to smooth transition between scenes.%\cite{GameEngineArchBook}\cite{GameProgPatternsBook}

@@

\begin{figure*}[b]
    \centering
    \includegraphics[width=0.99\linewidth]{Other/Figures/MethodologyDiagramCropped.pdf}
    \caption{Shows a flow diagram of the methodology proposed to define the level of non-determinism of a physics engine.}

    \label{method_diagram}
\end{figure*}

% ======= Sources of non-determinism
\subsection{Sources of non-determinism} \label{s:nondeterminisimSources}
\noindent After establishing an understanding of how physics engines work, the focus will now be on what causes these engines to be non-deterministic. 
The main reasons that we think are causes of these engines' non-determinism are discussed below. \\\\
%
\noindent\underline{\textit{Floating Points:}}
A generic issue with computers are floating points precision. 
Various errors occur when doing arithmetic manipulations using floating points, especially when operating between large and small numbers due to rounding, memory limitations and so forth.\cite{FloatingPointArithmeticArticle}\cite{FloatingPointsBook}.

Translating to the usage of CAV simulations in physics engines, this could result in many precision issues; for instance consider a simulation where the world is infinite and progressively  generated in the physics engine. 
After a few hundred meters, precision issues will start to occur due to the coordinates used in arithmetic cacluations getting bigger, and will get progressively worse the further from the origin the simulation gets. 
Such reasons would often make one rapidly jump to conclude that floating points are the cause of non-determinism of these engines. 
However, it is important to note that these precision issues should be referred to as incorrecctness and not non-determinisim. 
Since hardware is deterministic, then the same hardware should produce the same exact output for the same inputted data i.e. even if the output is incorrect due to floating point precision it should always be the same, given the same hardware.\\\\ 
%
% A generic issue with computers are floating points precision. 
%Various errors occur when doing arithmetic manipulations using floating points, especially when operating between large and small numbers due to rounding, memory limitations and so forth.\cite{FloatingPointArithmeticArticle}\cite{FloatingPointsBook}. 
%Translating to the usage of CAV simulations for V\&V in physics engines, this could result in many precision issues resulting in complex non deterministic behaviours.
%
% For instance, consider a simulation where the world is infinite and progressively  generated in the physics engine. 
% After a few hundred meters, precision issues will start to occur, and will get progressively worse the further from the origin the simulation gets. 
% This is due to the values of the coordinates being used in calculations are getting bigger and hence will not cope with small values that are critical for V\&V testing purposes. 
% A solution to this problem is perhaps every time the simulations moves 100 meters away from the origin the entire world will move by 100m in the opposite direction. Thus, avoiding getting into floating point issues.\cite{FloatingPointArithmeticArticle}
%
% There are many scenarios that could be addressed, and solutions to them will entirely depend on the characteristics of that specific scenario. There is no generalised solution for the floating problem, regardless of how good computers get there will always be a limitation. However, floating points as they stand could just be sufficient for V\&V testing purposes. Whenever they are not, smart solutions can be found to accommodate for these limitations.\\\\
%
\noindent\underline{\textit{Scheduling of threads:}} Scheduling of threads is believed to be the main reason for non-deterministic behaviour of game engines. Given the same hardware the output should always be the same, unless scheduling of the threads change i.e. the utilisation of hardware is manipulated differently each time. This would then result in non-deterministic behaviours especailly if it is combined with the floating point problem discussed earlier. 
To solve the scheduling of threads problem, one would have to control all of the threads during runtime, which is very involved. To achieve that one would need to replace or control the whole run-time system to allocate tasks to the same threads in the same order. This is beyond the sope of our work and will not be covered in this paper.\\\\
%
\noindent\underline{\textit{Physics and Rendering clocks:}}
The inherent way of how these physics engines operate (i.e.game loops) causes the engine to be non-deterministic. 
A follow up from section \ref{GameLoopSection}, these engines use a variable frame rate, which is good for hardware scalability but creates a challenge for the physics engine which works best with small fixed time steps. 
The problem of having a variable frame rate could also contribute to the non-deterministic behaviour of these engines, especially if put together with the scheduling of threads problem. That is to say, if the frame rate is allowed to vary then it has to be controlled to give the same variablity through out a given test in order for the test results to be repeatable. This issue might be of negligible significance if the frame rate is consistent and the time steps are small enough. \\\\
%
\noindent\underline{\textit{Navigation meshes and AI:}}
Game engines' often use navigation meshes for AI path planning. Some CAV simulator developers claim that the built-in physics engines' AI is non-deterministic\cite{CARLABenchmark}. 
The validity of this depends on what are the built in AI algorithms being used by the engine. 
Most of them tend to use the A* algorithm\cite{AStarBook}, which is an algorithm with deterministic behaviour if the environment is deterministic\cite{AirsimUnrealArticle}\cite{UnrealAIDocumentation}. 
Therefore, the determinism of the AI depends on the determinism of the engine and how its navigation meshes are created/modelled, their shape, granularity etc. It is interesting to note that changes can occur to meshes every time the simulation is loaded or if dynamically loading these navigation meshes as the simulation proceeds. This non-deterministic behaviour can be also resorted to the different scheduling of threads, hence giving variations in meshes everytime they are loaded.

### 2019-12-19 paper/3_Methodology.tex
Added discussed points and edited structure a bit, need to ask greg about some of the changes he did

\section{Methodology} \label{method}
\noindent A method for determining the level of simulation variance of an engine is presented in this section and the working flow of the method is shown in Figure~\ref{method_diagram}. Each of the steps in the method is elaborated in the following subsections.

\subsection{Design experiment}
The experiment should be a simple scenario that can be easily repeated with the same initial conditions. Logs should be used to record the actor positions at fixed time intervals throughout the simulation. For each actor the variance in trajectory can be determined and the deviation from the mean, $S_t$, of the trajectory monitored at each time step. Averages over time and for different actors can be used to condense results down to single figures which is done for some results.

We use game engine actors to simulate the AV and pedestrians. Actors can either follow a series of trajectory waypoints or can plan a route to a further destination using some path planning and obstacle avoidance algorithms, e.g. A* search algorithm. 
% 
% Resource - Realism
To simulate the demands of realistic rendering and complex physics calculations in a simple scenario, stress routines are used to artificially occupy system resources. 
% Collision callbacks
Further to this, the experiment should be designed to create collision callbacks to the physics engine by allowing actors to collide during the scenario. 
%
The user should also design the experiment to monitor and record system CPU and GPU utilisation. Additionally, if the user is interested in real-time assessment of the simulation then FPS should also be monitored and logged.
%
Where possible a fixed $dt$ should be used and any random numbers used should be seed controlled.

% \noindent When designing the experiment(s) one needs to bare in mind the following two questions: i) Is the engine deterministic? ii) If not, how does its simulation variance vary? In simulation, actors can either follow a series of trajectory way points or they get given the destination and their AI path plans to that destination. Thus, given that an engine has a high level of simulation variance, then one needs to set up an experiment that would stress the engine in order to determine how its simulation variance vary. 
% This is best done by generating collision callbacks, because that is when engines' do a lot of calculations to determine the response after the collision. Another approach, if AI is used for path planning instead of just following predefined trajectory way points, is to introduce a decision point for the AI.\\\\
It is crucial to also make sure that any randomisation factors are eliminated in the experiment.

\subsection{Internal settings}
\noindent There are several error and physics collision internal settings in physics engines that can be tweaked to enhance the simulation variance of the engine. \\\\
\noindent Increasing the physics time-step calculations, which is increasing the number of calculations per unit time, would improve the level of simulation variance since the physics sequence will be more finely defined; but on the other hand this would be more computationally expensive. However, in such applications of V\&V of CAV simulations, the computational cost should be of less concern compared to the importance of repeatability of tests. \\\\
\noindent If the experiment setup includes the usage of the physics engine's AI then altering the navigation mesh settings in the engine, like increasing the granularity of the mesh or changing the mesh type; can improve the level of simulation variance as a result of allowing the AI to navigate in a more well defined space.  \\\\
\noindent Disabling rendering or running in headless mode is a way of attempting to entirely decouple the rendering engine from the physics engine, which theoretically should affect positively the repeatablility of tests. 
The downside of most physics engines is that they do not provide a headless mode in the editor setup, thus this could make running in headless mode a real challenge for verification engineers.  \\\\
\noindent Running in editor or deployed mode can sometimes cause massive differences in performance, with deployed mode being worse, mostly due to the way settings get packaged when in deployed mode.

\begin{table*}[b]
\centering
\begin{tabular}{clclcc}
\toprule
Test ID & Test description (See Fig \ref{Test_a} and \ref{Test_b}) & Collision/Intersection & Collision Type & Look ahead distance (m) & No. of repeats \\ \midrule
1       & Two vehicles driving                   & No  & N/A & 2 & 1000 \\
2       & Two vehicles driving                   & Yes & Vehicle and Vehicle & 2 & 1000 \\
3       & Two vehicles driving and a pedestrian  & No  & N/A & 2 & 1000 \\
4       & Two vehicles driving and a pedestrian  & Yes & Vehicle and Pedestrian & 2 & 1000 \\
5       & Two pedestrians                        & No  & N/A & 2 & 1000 \\
6       & Two pedestrians                        & Yes & Pedestrian and Pedestrian & 0.4 & 1000 \\
7       & Two pedestrians                        & Yes & Pedestrian and Pedestrian & 2 & 1000 \\
8       & Two pedestrians                        & Yes & Pedestrian and Pedestrian & 20 & 1000 \\
\bottomrule
\end{tabular}
\caption{Set of experiments}
\label{TableOfExperiments}
\end{table*}

\subsection{External settings}
\noindent Running other programs in the background, like having a web browser open, or running the physics engine in the background while performing other tasks on the same machine does also have an effect; because it alters the CPU and GPU utilisation of tasks.\\\\ 
\noindent Trying to run stress experiments on computers by running other programs makes it difficult to control experiments since the utilisation would be inconsistent, nonetheless, there exists dedicated stress utilisation programs which provide consistent CPU and GPU stressing. Note that during running these tests computers should be left alone and not to be used for any other purposes apart from running the experiments, else the CPU + GPU stressing will lose its consistency through out a given test.

@@

\subsection{Output and analysis}
\noindent Once the experiments are run, the logged data is post processed to find the variance in the logged data between the different runs. Note that bifurcation effects can cause a jump in the variance, so it is worth plotting the different logged paths in order to determine if there are any bifurcation effects.\\\\
This whole process is repeated for various internal and external settings (if needed). 
Then various plots can be created to draw the engine's simulation variance. 
Thereon, it would be up to the verification engineers to determine where they want to define the line below which the simulation variance would be acceptable, and thus know at which levels of computational utilisation they can guarantee to run repeatable experiments.

>>>>

\section{Methodology} \label{s:methodology}
\noindent A method for determining the level of simulation variance of an engine is presented in this section and the working flow of the method is shown in Figure~\ref{method_diagram}. Each of the steps in the method is elaborated in the following subsections.
%
\subsection{Design experiment}
The experiment should be a simple scenario that can be easily repeated with the same initial conditions. Logs should be used to record the actor positions at fixed time intervals throughout the simulation. For each actor the variance in trajectory can be determined and the deviation from the mean, $S_t$, of the trajectory monitored at each time step. Averages over time and for different actors can be used to condense results down to single figures.

We use game engine actors to simulate the AV and pedestrians. Actors can either follow a series of trajectory waypoints or can plan a route to a further destination using some path planning and obstacle avoidance algorithms, e.g. A* search algorithm. 
% 
% Resource - Realism
To simulate the demands of realistic rendering and complex physics calculations in a simple scenario, stress routines are used to artificially occupy system resources. 
% Collision callbacks
Further to this, the experiment should be designed to create collision callbacks to the physics engine by allowing actors to collide during the scenario. 
%
The user should also design the experiment to monitor and record system CPU and GPU utilisation. Additionally, if the user is interested in real-time assessment of the simulation then FPS should also be monitored and logged.
%
Where possible a fixed $dt$ should be used and any random numbers used should be seed controlled.
% \noindent When designing the experiment(s) one needs to bare in mind the following two questions: i) Is the engine deterministic? ii) If not, how does its simulation variance vary? In simulation, actors can either follow a series of trajectory way points or they get given the destination and their AI path plans to that destination. Thus, given that an engine has a high level of simulation variance, then one needs to set up an experiment that would stress the engine in order to determine how its simulation variance vary. 
% This is best done by generating collision callbacks, because that is when engines' do a lot of calculations to determine the response after the collision. Another approach, if AI is used for path planning instead of just following predefined trajectory way points, is to introduce a decision point for the AI.\\\\
It is crucial to also make sure that any randomisation factors are eliminated in the experiment.
%
\subsection{Internal settings}
There are several error and physics collision internal settings in physics engines that can be tweaked to enhance the simulation variance of the engine.

Increasing the physics time-step calculations, which is increasing the number of calculations per unit time, would improve the level of simulation variance since the physics sequence will be more finely defined; but on the other hand this would be more computationally expensive. However, in such applications of V\&V of CAV simulations, the computational cost should be of less concern compared to the importance of repeatability of tests.

If the experiment setup includes the usage of the physics engine's AI then altering the navigation mesh settings in the engine, like increasing the granularity of the mesh or changing the mesh type; can improve the level of simulation variance as a result of allowing the AI to navigate in a more well defined space.

Disabling rendering or running in headless mode is a way of attempting to entirely decouple the rendering engine from the physics engine, which theoretically should affect positively the repeatablility of tests. 
The downside of most physics engines is that they do not provide a headless mode in the editor setup, thus this could make running in headless mode a real challenge for verification engineers.

Running in editor or deployed mode can sometimes cause massive differences in performance, with deployed mode being worse, mostly due to the way settings get packaged when in deployed mode.

\begin{table*}[b]
\centering
\begin{tabular}{clclcc}
\toprule
Test ID & Test description (See Figures~\ref{Test_a} and \ref{Test_b}) & Collision/Intersection & Collision Type & Look ahead distance (m) & No. of repeats \\ \midrule
1       & Two vehicles driving                   & No  & N/A & 2 & 1000 \\
2       & Two vehicles driving                   & Yes & Vehicle and Vehicle & 2 & 1000 \\
3       & Two vehicles driving and a pedestrian  & No  & N/A & 2 & 1000 \\
4       & Two vehicles driving and a pedestrian  & Yes & Vehicle and Pedestrian & 2 & 1000 \\
5       & Two pedestrians                        & No  & N/A & 2 & 1000 \\
6       & Two pedestrians                        & Yes & Pedestrian and Pedestrian & 0.4 & 1000 \\
7       & Two pedestrians                        & Yes & Pedestrian and Pedestrian & 2 & 1000 \\
8       & Two pedestrians                        & Yes & Pedestrian and Pedestrian & 20 & 1000 \\
\bottomrule
\end{tabular}
\caption{Set of experiments}
\label{TableOfExperiments}
\end{table*}

\subsection{External settings}
Running other programs in the background, like having a web browser open, or running the physics engine in the background while performing other tasks on the same machine does also have an effect; because it alters the CPU and GPU utilisation of tasks.

Trying to run stress experiments on computers by running other programs makes it difficult to control experiments since the utilisation would be inconsistent, nonetheless, there exists dedicated stress utilisation programs which provide consistent CPU and GPU stressing. Note that during running these tests computers should be left alone and not to be used for any other purposes apart from running the experiments, else the CPU + GPU stressing will lose its consistency through out a given test.

@@

\subsection{Output and analysis}
Once the experiments are run, the logged data is post processed to find the variance in the logged data between the different runs. Note that bifurcation effects can cause a jump in the variance, so it is worth plotting the different logged paths in order to determine if there are any bifurcation effects.

This whole process is repeated for various internal and external settings (if needed). 
Then various plots can be created to draw the engine's simulation variance. 
Thereon, it would be up to the verification engineers to determine where they want to define the line below which the simulation variance would be acceptable, and thus know at which levels of computational utilisation they can guarantee to run repeatable experiments.
